---
output: github_document
bibliography: vignettes/MSbiblio.bib
csl: vignettes/aaps-pharmscitech.csl
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please only edit README.Rmd -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# msprog: reproducible assessment of disability course in MS 

<!-- badges: start -->
[![R-CMD-check](https://github.com/noemimontobbio/msprog/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/noemimontobbio/msprog/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`msprog` is an R package providing tools for exhaustive and reproducible analysis of disability course in multiple sclerosis (MS) from longitudinal data [@msprog2024].

Its core function, `MSprog()`, detects and characterises the evolution of an outcome measure (EDSS, NHPT, T25FW, SDMT; or any custom outcome measure) for one or more subjects, based on repeated assessments through time and on the dates of acute episodes (if any).

The package also provides two toy datasets for function testing:

* `toydata_visits`: artificially generated EDSS and SDMT assessments for four patients;
* `toydata_relapses`: artificially generated relapse onset dates associated with the patients in `toydata_visits`.

Please refer to the documentation for function usage (e.g. `?MSprog`) and data structure (e.g. `?toydata_visits`). The whole documentation can be found in the [reference manual (PDF)](msprog.pdf). Additionally, a detailed tutorial providing examples and best-practice tips is available as a [package vignette](#vignette): *Analysing disability course in MS*. 

<!-- The outcome computation can be run locally on any computer with R version $\geq$ 3.5.0 (see installation instructions below), or online via a user-friendly [web application](https://msprog.shinyapps.io/msprog/). -->

**If you use this package in your work, please cite it [as below](#citation)**.

For any questions, requests for new features, or bug reporting, please contact: noemi.montobbio\@unige.it.<br /> Any feedback is highly appreciated!

<a id="install"></a>

## Installation

You can install the development version of `msprog` from GitHub by running the command below.
If using RStudio, please make sure to restart the R session right before running the installation.

``` r
# install.packages("devtools") # if not already installed
devtools::install_github("noemimontobbio/msprog", build_vignettes=TRUE)
```

## Usage

`MSprog()` detects the events sequentially by scanning the outcome values in chronological order, and classifies worsening events as relapse-associated or relapse-independent based on their relative timing with respect to the relapses [@lublin2014; @kappos2018; @silent2019].

Several qualitative and quantitative options for event detection are given as arguments that can be set by the user and reported as a complement to the results to ensure reproducibility. These include the baseline scheme (fixed or roving), the events to be detected (first or multiple, worsening and/or improvement), the length of the relapse influence window and of the event confirmation period(s) with the relative tolerance.

The example below illustrates the function's usage and output:

```{r}
library(msprog)

# Load toy data
data(toydata_visits)
data(toydata_relapses)

# Compute disability course
output <- MSprog(toydata_visits,                                      # provide data on visits
                 subj_col='id', value_col='EDSS', date_col='date',    # specify column names
                 outcome='edss',                                      # specify outcome type
                 event='multiple', baseline='roving',                 # modify default options
                 conf_tol_days=0, conf_unbounded_right=T,             # modify default options
                 relapse=toydata_relapses)                            # provide data on relapses
```
The function prints out a concise report of the results, and of the options used to obtain them. Complete results are stored in an object of class `MSprogOutput` with the following attributes.

1. `event_count`: a `data.frame` containing the event sequence detected for each subject, and the counts for each event type.
```{r, indent = "     "}
print(output$event_count)
```
    where: `event_sequence` specifies the order of the events; the other columns count the events of each type.
    
2. `results`: extended info on each event for all subjects.
```{r, indent="     "}
print(output$results, row.names=FALSE)
```
    where: `nevent` is the cumulative event count for each subject; `event_type` characterises the event; `time2event` is the number of days from start of follow-up to event; `bl2event` is the number of days from current baseline to event; `conf84` reports whether the event was confirmed over 84 days (12 weeks); `sust_days` is the number of days for which the event was sustained; `sust_last` reports whether the event was sustained until the last visit. 


Additionally, applying the `print` method to an object of class `MSprogOutput` prints out a short paragraph describing the complete set of criteria used to obtain the output, **to be reported to ensure complete reproducibility**:
```{r}
print(output)
```

<br />

<a id="vignette"></a>

### Vignettes
For detailed indications on usage and best practices, please refer to the package vignettes (make sure to enable `build_vignettes` during [installation](#install)). 
```{r, message=FALSE}
browseVignettes('msprog')
```

<a id="citation"></a>

### Citation

If you use the `msprog` package, please use the `citation()` function to obtain the correct reference:
```{r}
citation('msprog')
```
    
## References
