---
output: github_document
bibliography: MSbiblio.bib
csl: aaps-pharmscitech.csl
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# msprog: reproducible assessment of disability progression in MS

<!-- badges: start -->
<!-- badges: end -->

`msprog` provides tools for exhaustive and reproducible analysis of disability progression in multiple sclerosis (MS) from longitudinal data.

Its core function, `MSprog`, detects and characterises the progression events of an outcome measure (EDSS, NHPT, T25FW, or SDMT) for one or more subjects, based on repeated assessments through time and on the dates of acute episodes.

The package also provides two toy datasets:

* `toydata_visits`: artificially generated EDSS and SDMT assessments for four patients;
* `toydata_relapses`: artificially generated relapse dates associated with the patients in `toydata_visits`.



## Installation

You can install the development version of `msprog` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("noemimontobbio/msprog")
```

## Usage

`MSprog` detects the events sequentially by scanning the outcome values in chronological order, and classifies progression events as relapse-associated or relapse-independent based on their relative timing with respect to the relapses [@lublin2014; @kappos2018; @silent]. Several qualitative and quantitative options are given as arguments that can be set by the user and reported as a complement to the results to ensure reproducibility. These include the baseline scheme (fixed or roving), the events to be detected (first or multiple, progression and/or improvement), the length of the relapse influence window and of the event confirmation period(s) with the relative tolerance. 

```{r}
library(msprog)

data(toydata_visits)
data(toydata_relapses)

output <- MSprog(toydata_visits,                                   # insert data on visits
                 subj_col='id', value_col='EDSS', date_col='date', # specify column names
                 event='multiple', baseline='roving',              # modify default options on event detection
                 relapse=toydata_relapses)                         # insert data on relapses
```
The function prints out a concise report of the results, as well as **the specific set of options used to obtain them**.

Complete results are stored in two `data.frame` objects generated by the function call.

1. A summary table providing the event count for each subject and event type:
```{r, indent = "     "}
summary = output[[1]]
print(summary)
```
    where: `event_sequence` specifies the order of the events; the other columns count the events of each type.
    
2. Extended info on each event for all subjects:
```{r, indent = "     "}
results = output[[2]]
print(results, row.names = FALSE)
```
    where: `nevent` is the cumulative event count for each subject; `event_type` characterises the event; `time2event` is the number of days from baseline to event; `conf3` reports whether the event was confirmed; `sust_days` is the number of days for which the event was sustained *after confirmation*; `sust_last` reports whether the event was sustained until the last visit.
    
    
## References
