---
output: github_document
bibliography: vignettes/MSbiblio.bib
csl: vignettes/aaps-pharmscitech.csl
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# msprog: reproducible assessment of disability progression in MS

<!-- badges: start -->
[![R-CMD-check](https://github.com/noemimontobbio/msprog/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/noemimontobbio/msprog/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`msprog` provides tools for exhaustive and reproducible analysis of disability progression in multiple sclerosis (MS) from longitudinal data.

Its core function, `MSprog`, detects and characterises the progression events of an outcome measure (EDSS, NHPT, T25FW, or SDMT) for one or more subjects, based on repeated assessments through time and on the dates of acute episodes.

The package also provides two toy datasets for function testing:

* `toydata_visits`: artificially generated EDSS and SDMT assessments for four patients;
* `toydata_relapses`: artificially generated relapse dates associated with the patients in `toydata_visits`.

Please refer to the documentation for function usage (e.g. `?MSprog`) and data structure (e.g. `?toydata_visits`). A detailed tutorial is available as a package vignette: ***Computing MS progression from longitudinal data***.


## Installation

You can install the development version of `msprog` from [GitHub](https://github.com/noemimontobbio/msprog) with:

``` r
# install.packages("devtools")
devtools::install_github("noemimontobbio/msprog", build_vignettes=TRUE)
```

## Usage

`MSprog` detects the events sequentially by scanning the outcome values in chronological order, and classifies progression events as relapse-associated or relapse-independent based on their relative timing with respect to the relapses [@lublin2014; @kappos2018; @silent].

Several qualitative and quantitative options for event detection are given as arguments that can be set by the user and reported as a complement to the results to ensure reproducibility. These include the baseline scheme (fixed or roving), the events to be detected (first or multiple, progression and/or improvement), the length of the relapse influence window and of the event confirmation period(s) with the relative tolerance.

The example below illustrates the function's usage and outputs:

```{r}
library(msprog)

# Load toy data
data(toydata_visits)
data(toydata_relapses)

# Compute progression
output <- MSprog(toydata_visits,                                   # provide data on visits
                 subj_col='id', value_col='EDSS', date_col='date', # specify column names
                 outcome='edss',                                   # specify outcome type
                 event='multiple', baseline='roving',              # modify default options on event detection
                 relapse=toydata_relapses)                         # provide data on relapses
```
The function prints out a concise report of the results, as well as **the specific set of options used to obtain them**. Complete results are stored in two `data.frame` objects generated by the function call:

1. A summary table providing the event count for each subject and event type:
```{r, indent = "     "}
summary = output[[1]]
print(summary)
```
    where: `event_sequence` specifies the order of the events; the other columns count the events of each type.
    
2. Extended info on each event for all subjects:
```{r, indent = "     "}
results = output[[2]]
print(results, row.names = FALSE)
```
    where: `nevent` is the cumulative event count for each subject; `event_type` characterises the event; `time2event` is the number of days from start of follow-up to event; `bl2event` is the number of days from current baseline to event; `conf12` reports whether the event was confirmed at 12 weeks; `sust_days` is the number of days for which the event was sustained *after confirmation*; `sust_last` reports whether the event was sustained until the last visit.

<br />

For detailed indications on usage and best practices, please refer to the package vignettes.
```{r, message=FALSE}
browseVignettes('msprog')
```
    
## References
