---
title: "Time to event"
author: "Noemi Montobbio"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
bibliography: MSbiblio.bib
csl: aaps-pharmscitech.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Time to event}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

```{r wrap-hook, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = xfun::split_lines(x) #substring(x, 4, nchar(x)))
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste0(paste(x, collapse = '\n#> '), '\n')
  }
  hook_output(x, options)
})
```

This vignette demonstrates how to use the `msprog` package to analyse the disability course in multiple sclerosis (MS) in a time-to-event setting (e.g., to compute survival times for survival analysis).

For a more general introduction on `msprog` package usage for disability course assessment, please refer to the vignette *Analysing disability course in MS*. 

We shall use the toy datasets `toydata_visits` and `toydata_relapses` provided in `msprog`.

```{r}
library(msprog)
head(toydata_visits)
head(toydata_relapses)
```

<!-- ```{r} -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/MSprog.R') -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/event_testing.R') -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/relapse_indep_from_bounds.R') -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/msprog_methods.R') -->
<!-- load('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/data/toydata_visits.RData') -->
<!-- load('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/data/toydata_relapses.RData') -->
<!-- library(dplyr) -->
<!-- ``` -->

## Time to first disability worsening event

Given the relative rarity of disability worsening events in MS, it is quite common to use the **time to *first* confirmed disability worsening** as an endpoint of interest. This is the case especially in clinical trials, where follow-up periods are often limited to 2-3 years. Survival times may be computed using the following code:
```{r}
output_edss <- MSprog(toydata_visits,
                 subj_col='id', value_col='EDSS', date_col='date',
                 outcome='edss',
                 event='firstCDW',   # <--- only detect first CDW event
                 relapse=toydata_relapses,
                 verbose=0)
```

Note in particular that the `event` argument is set to `firstCDW` -- i.e., only the first confirmed disability worsening (CDW) event is detected. Many other arguments are not explicitly set: for those, the defaults are used (you can type `?MSprog` to access the documentation explaining how to specify each argument). Let's print out a description of the criteria used for the above computation, to make sure all the settings are as expected. 
```{r, linewidth=90}
print(output_edss)
```

To access the time-to-event data, we need to extract the `results` attribute from the function output.
```{r}
res <- output_edss$results
# print(res, row.names=FALSE)
DT::datatable(res, rownames=F,
              options = list(dom='t', scrollX=T, scrollY=F, paging = FALSE)
              )
```

If we were to conduct a survival analysis with censoring, we would be interested in the `'time2event'` column (time to CDW, in days) and in the `'nevent'` column (1 if the event occurred, 0 otherwise):
```{r}
survival_data <- res[c('id', 'time2event', 'nevent')]
# print(survival_data, row.names=FALSE)
DT::datatable(survival_data, rownames=F,
              options = list(dom='t', scrollX=T, scrollY=F, paging = FALSE)
              )
```
Note that, in a more general multiple-event setting, the `'nevent'` column would contain the count of events for each subject. Here, since we're stopping at the first event, this number can only be 0 (no event) or 1 (event).

The data is now ready to use for survival analysis. Here's an example code that one may use to plot Kaplan-Meier curves from `survival_data`:
```{r, eval=FALSE}
# library(dplyr)
# library(ggsurvfit)
survfit2(Surv(time2event, nevent) ~ 1, data=survival_data) %>%
  ggsurvfit() +
  labs(
    x = "time (days)",
    y = "survival probability"
  )
```

## Time to disability milestone

<!-- ```{r} -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/value_milestone.R') -->
<!-- library(dplyr) -->
<!-- ``` -->

Instead of studying disability worsening with respect to a baseline value, one can focus on the time taken to reach a specific disability milestone. This can be computed using `msprog::value_milestone()`.

The following code detects the time to EDSS $\geq$ 4.5 for all subjects in our toy data. Similar to `MSprog()`, we can set `verbose=2` to display progress info.
```{r}
vm <- value_milestone(toydata_visits, milestone=4.5,
                 subj_col='id', value_col='EDSS', date_col='date',
                 outcome='edss', relapse=toydata_relapses,
                 verbose=2)
```

For example, subject `2` reached EDSS=4.5 at their third visit, on date 2021-03-24:
```{r}
print(toydata_visits[toydata_visits$id==2, c('date', 'EDSS')], row.names=FALSE)
```
Subject `4` had EDSS=4.5 at their first visit, but it was not confirmed; the first confirmed value $\geq$ 4.5 was found at the fourth visit, on date 2022-07-19:
```{r}
print(toydata_visits[toydata_visits$id==4, c('date', 'EDSS')], row.names=FALSE)
```


The function returned the following `data.frame`:
```{r}
# print(vm)
DT::datatable(vm,
              options = list(dom='t', scrollX=T, scrollY=F, paging = FALSE)
              )
```
where: `'date'` contains the date of the first confirmed EDSS $\geq$ 4.5 (or last date of follow-up if milestone was not reached or not confirmed); `'EDSS'` contains the first EDSS value $\geq$ 4.5, if present, otherwise no value; `'time2event'` contains the time to reach EDSS $\geq$ 4.5 (or total follow-up length if not reached or not confirmed); `'observed'` indicates whether EDSS $\geq$ 4.5 was reached (1) or not (0).

Please refer to the documentation (by typing `?value_milestone`) for a complete illustration of each of the function arguments and their default values.

As before, for survival analysis we would need the `'time2event'` column (time to reach EDSS $\geq$ 4.5, in days) and the `'observed'` column (1 if the milestone was reached, 0 otherwise).

## Multiple events

In studies with longer follow-ups, it can be informative to analyse *recurrent events* with consecutive survival times.

The following code extracts multiple EDSS events from our toy data with a **roving baseline** scheme (i.e., the baseline is updated after each event to the first available confirmation visit).
```{r}
output <- MSprog(toydata_visits, 'id', 'EDSS', 'date', 'edss', relapse=toydata_relapses,
                event='multiple', baseline='roving',  # <--- detect multiple events with a roving baseline
                verbose=0)
survival_data <- output$results[c('id', 'nevent', 'time2event')]
# print(survival_data, row.names=FALSE)
DT::datatable(survival_data, rownames=F,
              options = list(dom='t', scrollX=T, scrollY=F, paging = FALSE)
              )
```
Note that the `event` argument is set to `multiple`, and that the results can contain more than one row for the same subject, with the `'nevent'` column keeping track of cumulative event count.

