---
title: "Computing MS progression from longitudinal data"
author: "Noemi Montobbio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: MSbiblio.bib
csl: aaps-pharmscitech.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Computing MS progression from longitudinal data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

This vignette illustrates how to use the `msprog` package to study the progression of disability in multiple sclerosis (MS) based on repeated assessments of an outcome measure (EDSS, NHPT, T25FW, or SDMT) through time, and on the dates of acute episodes (if present).

```{r}
library(msprog)
```

<!-- ```{r} -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/MSprog.R') -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/event_testing.R') -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/relapse_indep_from_bounds.R') -->
<!-- source('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/R/msprog_methods.R') -->
<!-- load('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/data/toydata_visits.RData') -->
<!-- load('/Users/nmontobbio/code/MSprog/R-MSprog/msprog/data/toydata_relapses.RData') -->
<!-- library(dplyr) -->
<!-- ``` -->

## Input data

The data must be organised in a `data.frame` object containing (at least) the following columns:

* Subject IDs;
* Visit dates;
* Outcome values.

The visits should be listed in chronological order (if they are not, `msprog::MSprog()` will sort them).

The `msprog` package provides a toy dataset `toydata_visits` with artificially generated EDSS and SDMT assessments for four patients.
```{r}
head(toydata_visits)
```

<br />

For relapsing-remitting MS patients, an additional `data.frame` object with the dates of relapses is needed to correctly assess progression and characterise progression events as relapse-associated or relapse-independent. The dataset should contain (at least) the following columns:

* Subject IDs;
* Visit dates.

A toy dataset with artificially generated relapse dates associated with the patients in `toydata_visits` is provided:
```{r}
head(toydata_relapses)
```

## Minimal example

Given data on visits and relapses in the form specified above, the `MSprog()` function analyses the disability progression for each subject for the outcome of interest. We can test it on the EDSS toy data:
```{r}
output_edss <- MSprog(toydata_visits, # data on visits
                 subj_col='id', value_col='EDSS', date_col='date', # specify column names
                 outcome='edss', # specify outcome type
                 relapse=toydata_relapses) # data on relapses
```
or on the SDMT toy data:
```{r}
output_sdmt <- MSprog(toydata_visits, # data on visits
                 subj_col='id', value_col='SDMT', date_col='date', # specify column names
                 outcome='sdmt', # specify outcome type
                 relapse=toydata_relapses) # data on relapses
```
The function prints concise info (the argument `verbose` can be used to control the amount of printed info), and outputs an object of class `MSprogOutput`. The following methods are available for extracting info from the output.

1. The `event_count` method generates a `data.frame` containing the event counts for each subject, and the event sequence (where relevant):
```{r, indent="     "}
print(event_count(output_edss))
```
    where: `event_sequence`[^1] specifies the order of the events; the other columns count the events of each kind (improvement; progression; relapse-associated worsening, RAW; progression independent of relapse activity, PIRA; and progression that could not be classified as either RAW of PIRA with the available information). See [@lublin2014; @kappos2018; @silent2019].

2. The `results` method generates extended info on each event for all subjects:
```{r, indent="     "}
print(results(output_edss), row.names=FALSE)
```
    where: `nevent` is the cumulative event count for each subject; `event_type` characterises the event; `total_fu` is the number of days from start to end of follow-up; `time2event` is the number of days from start of follow-up to event[^2]; `bl2event` is the number of days from current baseline to event; `conf12` reports whether the event was confirmed over 12 weeks; `sust_days` is the number of days for which the event was sustained; `sust_last` reports whether the event was sustained until the last visit.

3. The `criteria_text` method prints out a short paragraph describing the complete set of criteria used to obtain the output, to be reported to ensure complete reproducibility:
```{r, indent="     "}
criteria_text(output_edss)
```


[^1]: In this example, `event_sequence` can contain a maximum of one event, as we are not specifying the argument `event` - which defaults to `firstprog` (only detect the first progression event). For the same reason, `nevent` in `results(output)` can only be 0 or 1 in this example. In a multiple-event setting (`event=multiple`), more than one event per subject can be detected. See [the relevant section](#multiple).
[^2]: For subjects with no events, `time2event` is the total follow-up length. To omit these subjects, set `include_stable=FALSE` in `MSprog()`.


<br />

Several qualitative and quantitative options for computing the progression are given as optional arguments of `MSprog()` that can be set by the user. In order to ensure reproducibility, the results should always be complemented by the set of criteria used to obtain them. In the following sections we will go into more detail about usage and best practices for the different options. Please refer to the documentation (by typing `?MSprog`) for a complete illustration of all function arguments.

## Valid changes in the outcome measure

The `MSprog()` function detects the events sequentially by scanning the outcome values in chronological order, and each value is tested for its difference from the current reference value. The `compute_delta()` function returns the minimum difference $\delta$ in the chosen outcome measure that is accepted as a valid change from a given reference value $x$ (value at baseline). The threshold $\delta$ is computed differently based on the test under analysis (see [@lorscheider2016; @bosma2010; @kalinowski2022; @strober2019]):

* Expanded Disability Status Scale (EDSS): $\delta(x)=\begin{cases} 1.5 \quad \text{ if } x=0\\1 \quad\;\;\; \text{ if } 0 < x \leq 5\\0.5 \quad \text{ if } 5<x\leq 10\end{cases}$;

* Nine-Hole Peg Test (NHPT), for either the dominant or the non-dominant hand: 20% of reference $x$;

* Timed 25-Foot Walk (T25FW): 20% of reference $x$;

* Symbol Digit Modalities Test (SDMT): either 3 points or 10% of reference $x$.


For example:
```{r}
print(compute_delta(4)) # default outcome measure is 'edss'
print(compute_delta(10, outcome='t25fw'))
```

The `compute_delta()` function is used as default `delta_fun` in the `MSprog()` function to assess progression. In this case, the threshold $\delta$ used to assess progression is computed based on the argument `outcome` given to `MSprog()`. For instance, calling:
```{r}
output <- MSprog(toydata_visits,
                 subj_col='id', value_col='SDMT', date_col='date',
                 outcome='sdmt', # ---> outcome type influences delta computation <---
                 relapse=toydata_relapses, verbose=0) 
```
disability progression will be defined as a worsening of either 3 SDMT points or 10% of the reference SDMT value. Alternatively, a custom `delta_fun` can be provided by the user to support different outcome measures, or to specify different rules for the supported ones. To change the minimum $\delta$ for SDMT to, say, "either 4 points or 20% of the reference value", we would define:
```{r}
my_sdmt_delta <- function(x) {min(unlist(c(x/5, 4)))}
print(my_sdmt_delta(50)) # my delta
print(compute_delta(50, outcome='sdmt')) # default delta
```
and set `delta_fun=my_sdmt_delta` in `MSprog()` when computing the progression.

## Baseline scheme and multiple events {#multiple}

The baseline scheme can be set by using the `baseline` argument in `MSprog()`. The following alternative baseline schemes can be adopted:

* Fixed baseline (`baseline='fixed'`, default): the reference value is set to be the first valid outcome value.
* Improvement-based roving baseline (`baseline='roving_impr'`): the reference value is initially set as the first valid outcome value, then updated after each confirmed improvement to the next valid outcome value. This scheme is suitable for a first-progression setting to discard fluctuations around the baseline [@muller2023]. The re-baseline procedure can be made finer by setting `sub_threshold=TRUE` in `MSprog()`: this moves the reference value after *any* confirmed improvement, even if the difference from the current reference is smaller than the minimum $\delta$.
* Roving baseline (`baseline='roving'`): the reference value is initially set as the first valid outcome value, then updated after each event to the last confirmed outcome value (out of relapse influence). **This scheme is recommended in a "multiple events" setting** [@kappos2018] (see example below). The re-baseline procedure can be made finer by setting `sub_threshold=TRUE` in `MSprog()`: this moves the reference value after *any* confirmed change, even if the difference from the current reference is smaller than the minimum $\delta$.

The argument `event` allows to specify which events to detect. By default, it is set to `firstprog` (only detect the first progression event). It can be set to `multiple` to sequentially detect all progression (or improvement) events.

For example, extracting multiple EDSS events for subject `4` from `toydata_visits` with a fixed baseline would result in the following.
```{r}
print(toydata_visits[toydata_visits$id==4, c('date', 'EDSS')], row.names=FALSE) # EDSS visits
output <- MSprog(toydata_visits, 'id', 'EDSS', 'date', 'edss', subjects=4,
                relapse=toydata_relapses, conf_weeks=12, event='multiple', baseline='fixed', 
                include_dates=TRUE, verbose=0)
print(results(output), row.names=FALSE) # results
```

Since the reference value was kept fixed at the first visit (EDSS=4.5), the EDSS progression at visit 4 (EDSS=5) was not detected. On the other hand, with a roving baseline scheme, the baseline is moved to visit 3 after the confirmed improvement and the progression event is correctly detected:
```{r}
output <- MSprog(toydata_visits, 'id', 'EDSS', 'date', 'edss', relapse=toydata_relapses, subjects=4,
                conf_weeks=12, event='multiple', baseline='roving', 
                include_dates=TRUE, verbose=0)
print(results(output), row.names=FALSE) # results
```

<br />

Finally, on top of the chosen baseline scheme, *post-relapse re-baseline* @kappos2020 can be applied by setting `relapse_rebl=TRUE` in `MSprog()`. If this is enabled, outcome values are scanned once more from the beginning by resetting the baseline after each relapse (first visit out of relapse influence) to classify potential events left undefined as PIRA.


## Confirmation period

An event is only validated if it is *confirmed*, that is, if the value change from baseline is maintained **up to** a subsequent visit falling within a pre-specified confirmation period [@ontaneda2017]. The chosen confirmation period depends on the type of study and on the frequency of visits, and can be set in `MSprog()` by using the argument `conf_weeks`. If multiple values are specified (e.g., `conf_weeks=c(12,24)`), events are retained if confirmed by at least a visit falling within one of the specified periods (here, 12 or 24 weeks $\pm$ `conf_tol_days`)[^3]. The results table will report whether an event was confirmed in each of the specified periods.

[^3]: An event is only confirmed if the value change from baseline is maintained **at all visits up to the confirmation visit**. So an event can only be "confirmed over 24 weeks" and *not* "confirmed over 12 weeks" if there are no valid confirmation visits falling within the 12-week window.


Let's look at subject `2` from `toydata_visits`:
```{r}
print(toydata_visits[toydata_visits$id==2,c('date', 'EDSS')], row.names=FALSE) # EDSS visits
print(toydata_relapses[toydata_relapses$id==2,], row.names=FALSE) # relapses
```

The following code detects 12- or 24-week confirmed events for subject `2`.
```{r}
output <- MSprog(toydata_visits, 'id', 'EDSS', 'date', 'edss', 
                 relapse=toydata_relapses, subjects=2,
                 event='multiple', baseline='roving', 
                 relapse_indep=relapse_indep_from_bounds(0,NULL,NULL,NULL,NULL,0),
                 conf_weeks=c(12,24), verbose=0)
print(results(output), row.names=FALSE) # results
```
The `relapse_indep` argument specifies the relapse-free intervals for a progression event to be classified as PIRA[^4]. In this case, a high-specificity definition of PIRA is used [@muller2023] that requires an absence of relapses between baseline and confirmation. The resulting events are a 12-week-confirmed RAW and a 12-week-confirmed PIRA. The RAW event was not confirmed over 24 weeks (`conf24` is `0`). The PIRA event was also confirmed over 24 weeks (`conf24` is `1`). However, since a relapse occurred before the 24-week confirmation, the event cannot be classified as a 24-week-confirmed PIRA (`PIRA_conf24` is `0`) but only as a 24-week-confirmed progression.

[^4]: See `relapse_indep_from_bounds()` function documentation, and the vignette *Assessing progression independent of relapse activity (PIRA) in MS*.

The RAW event found for subject `2` constitutes a *transient* accumulation of disability. Such events can be excluded from the `MSprog()` output by requiring that each event be sustained for at least a certain amount of time, specified by argument `require_sust_weeks`. For instance:
```{r}
output <- MSprog(toydata_visits, 'id', 'EDSS', 'date', 'edss', 
                relapse=toydata_relapses, subjects=2,
                event='multiple', baseline='roving', 
                relapse_indep=relapse_indep_from_bounds(0,NULL,NULL,NULL,NULL,0),
                conf_weeks=c(12,24), require_sust_weeks=48, verbose=0)
print(results(output), row.names=FALSE) # results
```

In this context, as the transient EDSS accumulation was not classified as an event, it did not trigger a re-baseline according to the roving baseline scheme. As a consequence, the PIRA event is classified as an undefined progression due to the presence of a relapse between baseline and confirmation. This can be handled by enabling *post-relapse re-baseline* (`relapse_rebl=TRUE`) to force a re-baseline after each relapse:
```{r}
output <- MSprog(toydata_visits, 'id', 'EDSS', 'date', 'edss', 
                relapse=toydata_relapses, subjects=2,
                event='multiple', baseline='roving', 
                relapse_indep=relapse_indep_from_bounds(0,NULL,NULL,NULL,NULL,0),
                relapse_rebl=TRUE,
                conf_weeks=c(12,24), require_sust_weeks=48, verbose=0)
print(results(output), row.names=FALSE) # results
```
The event is now correctly classified as PIRA.

A more detailed report of the event detection process in each of the three cases examined can be visualized by re-running the above code snippets with `verbose=2`, see next section.  

<br />

Finally, the tolerance for the confirmation visit date can be set using the argument `conf_tol_days`. If a single number is specified (e.g., `conf_tol_days=45`), a symmetric tolerance interval is used: if the confirmation period is, say, 12 weeks, any visit within $[12\text{weeks} - 45\text{days}, 12\text{weeks} + 45\text{days}]$ will be a valid confirmation visit. Different tolerance on the left and on the right can be set by specifying two values (e.g., `conf_tol_days=c(30, 365)` will generate a window $[12\text{weeks} - 30\text{days}, 12\text{weeks} + 365\text{days}]$). Further, the argument `conf_unbounded_right` allows to consider as valid any visit *after* a certain amount of time (e.g., `conf_tol_days=45` with `conf_unbounded_right=TRUE` will result in the window $[12\text{weeks} - 30\text{days}, +\inf]$).



## Printing progress info

The `MSprog()` function scans the outcome values of each subject in chronological order to detect the events. It is possible to visualize an extended log of the ongoing computations by setting `verbose=2`. See the example below.

```{r}
output <- MSprog(toydata_visits, 
                 subj_col='id', value_col='EDSS', date_col='date',
                 outcome='edss',
                 event='multiple', baseline='roving', 
                 relapse=toydata_relapses, verbose=2)
```

## References




